---
title: "Day 1 - Introduction to R and ggplot2"
author: "Amy Henry"
date: "9/23/2021"
output: html_document
---

This is where I write my normal text. 

Code needs to go in chunks, like this: 
```{r}

```

To run a line of R code (aka, send it to the Console), you can:
- Click the "Run" button above
- Press Command + Enter while the cursor is on the line (or highlighted) 

To run a whole chunk:
- click the green "Play" button on the right side of the chunk header. 

Command + Alt + I generates a new "chunk". Try it here! 

```{r}

```

When assigning a value to an object, R does not print anything. You can force R to print the value by using parentheses or by typing the object name:
```{r}
5
x <- 5
(x <- 5)
x
```


```{r}
# objects can store more than one value in a one dimensional vector
my_vector <- c(1,4,5,6,8)
```

Whole dataframes can also be assigned to objects. MPG is one that comes built in with the tidyverse package and describes specs of different models of car. 

Remember, packages are bundles of code and functions that other people have written and shared with us to use. They save us a lot of work, but need to be loaded fresh every time you open R. 

tidyverse is a very popular package that we will use functions from a lot, and is actually a collection of packages. One of the packages in it is ggplot2, which makes it very easy to make very nice looking plots. 
```{r}
library("tidyverse")
mpg

iris

#select one column
mpg$model

mpg$model[119]
```

##################################################3
## But come on, there's time for all that boring stuff later, you said we could make pretty graphs!

The fivethirtyeight package contains datasets written about on the FiveThirtyEight blog. We're going to play with a dataset of popular Halloween candies, their characteristics and how popular they are, as assessed in survey data. 

```{r}
library("fivethirtyeight")
```

What are some ways we can look at this dataset?
```{r}
candy_rankings
```

```{r}
head(candy_rankings)
```

```{r}
glimpse(candy_rankings)
```

```{r}
View(candy_rankings)
```

## GGPLOTS: The Basic format

There are three really basic components you need for any ggplot - the dataset, the variables you want on each axis, and a geom to tell it what kind of plot to make. Examples of geoms: geom_bar(), geom_histogram(), geom_boxplot(), geom_line() - all pretty intuitive about what they'll make! These pieces are layered into the plot with + signs. 

What it looks like when you put those things together
ggplot(<dataset>) + 
  aes(x = <variable you want on the x-axis from dataset>, y = <ditto for y>) + 
  geom_<typeofplot>(<modifiers of that plot>)
```{r}
ggplot(candy_rankings) # this line makes the space to put the plot, the background
```

```{r}
ggplot(candy_rankings) + 
  aes(x = chocolate) # this says what to put on the axes
```

```{r}
ggplot(candy_rankings) +  
  aes(x = chocolate) + 
  geom_bar() # this says what kind of plot to make, in this case, a barplot that automatically did counts of the observations of which candies were chocolate or not chocolate. 
```

ggplots can be defined as an object and recalled later too. This is useful when you are adding more geoms and trying things out, or when you're compiling the plots into a multiple plot figure or exporting them. 

```{r}
chocolate_bar <- ggplot(candy_rankings) +  
  aes(x = chocolate) + 
  geom_bar() 

chocolate_bar
```

Challenge: 
Plot a different variable, without copying and pasting the script you wrote before. 



#Okay, we've got barplots nailed. What other kinds of plots might be interesting? 

For example, here we've plotted the win percentage (how often people ranked it in their favorites) against sugar percent, to ask the question, do sweeter candies win more often?

We're using the geom_point() function to plot this as a scatter plot. 

```{r}
sweetness <- ggplot(candy_rankings) + 
  aes(x = sugarpercent, y = winpercent) + 
  geom_point()
```

To add a best-fit regression line, we can use the geom_smooth() function. 

```{r}
sweetness <- ggplot(candy_rankings) + 
  aes(x = sugarpercent, y = winpercent) + 
  geom_point()+
  geom_smooth(method = "lm")

#notice how you can add the layer to the already created object
sweetness + 
  geom_smooth(method = "lm")
```


# What about a boxplot? 

```{r}
sweetness <- ggplot(candy_rankings) + 
  aes(x = chocolate, y = sugarpercent) + 
  geom_boxplot()
sweetness
```

## Reference for which geoms/stats to use for which kinds of plots: 
Check out your ggplot2 Cheatsheet! This has a nice visual reference to what geom to use to make what kind of plot. 

More detail here, to not worry about yet. 
https://ggplot2.tidyverse.org/reference/

## Try making a histogram! Play around with some of the variables and plot types from these references. 

```{r}
sweethist <- ggplot(candy_rankings) + 
  aes(x = sugarpercent) + 
  geom_histogram()
sweethist
```


### Tests: 

1.  Are fruity candies cheaper? 
- Can you make a boxplot that shows the price percent depending on fruity?

2. Are sweeter candies more expensive? Do the candies with higher sugarpercent have higher winpercent? Plot this relationship.

3. What can you do with your own data?


PRETTY TIME! 

## Modifying colors
Reference: 
bit.ly/colors-r

You can change the color in the geom argument. Test out how it works on different geoms!
```{r}
ggplot(candy_rankings) +  
  aes(x = chocolate) + 
  geom_bar(color = 'darkblue', fill = 'coral') 
```


If you want the color to reflect another variable, for example, if there is fruit in the candy too, then you can assign that under aes(). 
```{r}
ggplot(candy_rankings) +  
  aes(x = chocolate, fill = fruity) + 
  geom_bar(position = position_dodge()) 
```

## Challenge question: what proportion of candies have BOTH chocolate and caramel? 





## Shapes
Working on a scatterplot and want the points to be different shapes? Easy too! 

```{r}
ggplot(candy_rankings) +
  aes(x = pricepercent,
      y = winpercent) +
  geom_point(shape = 2)
```

Similarly, you can change shape by some other variable. 
```{r}
ggplot(candy_rankings) +
  aes(x = pricepercent,
      y = winpercent, 
      shape = chocolate) +
  geom_point()
```

## Labels

Pretty self-explanatory here. labs() are added separately from the other plot parameters.
```{r}
ggplot(candy_rankings) +
  aes(x = sugarpercent, 
      y = winpercent) +
  geom_point() +
  labs(x = "Sugar Percentage",
       y = "Win Percentage",
       title = "Relationship between the dependent and independent variable",
       subtitle = "for Halloween candies",
       caption = "Based on an online Survey by FiveThirtyEight")

```


## X and Y axis limits

```{r}
ggplot(candy_rankings) +
  aes(x = sugarpercent, 
      y = winpercent) +
  geom_point() +
  xlim(0, 1) +
  ylim(0, 100)
```

## Examples of other types of plots

###Barplot

```{r}
ggplot(candy_rankings) +
  aes(x = chocolate) + 
  geom_bar()
```

## Histogram

```{r}
ggplot(candy_rankings) + 
  aes(x = winpercent) +
  geom_histogram()
```

### Density Plot

```{r}
ggplot(candy_rankings) + 
  aes(x = winpercent) +
  geom_density()
```


### Dot plot
```{r}
ggplot(candy_rankings) + 
  aes(x = winpercent) +
  geom_dotplot()
```
Note the error message - try adding the argument "binwidth = 3" to the geom_dotplot() line. 

### Boxplot 

```{r}
ggplot(candy_rankings) +
  aes(x = chocolate, 
      y = winpercent) +
  geom_boxplot()
```

### Scatter pLot

```{r}
ggplot(candy_rankings) +
  aes(x = pricepercent, 
      y = winpercent) +
  geom_point()
```

### Examples of multiple layered plots

```{r}
ggplot(candy_rankings) +
  aes(x = pricepercent, 
      y = winpercent) +
  geom_point() +
  geom_smooth()
```

With a line fitted using the lm() function instead of a smoothed line. Test out what happens when you set "se = TRUE". 
```{r}
ggplot(candy_rankings) +
  aes(x = pricepercent, 
      y = winpercent) +
  geom_point() +
  geom_smooth(method = "lm", 
              se = FALSE)
```

Multiple plots in multiple panes - 
what happens when you move the . and ~ next to chocolate, like in the commented out line? 
```{r}
ggplot(candy_rankings) +
  aes(x = sugarpercent,
      y = winpercent) +
  geom_point() +
  facet_grid(.~chocolate)
  #facet_grid(chocolate~.)
```



Done already? 

Practice plotting some more with: 
```{r}
drinks # how much of different kinds of booze do people in different countries drink?

fandango # do different movie rating systems agree with each other?

flying # How do people behave (or feel about how other people behave) on airplanes?
```


