---
title: "Data Types Demo"
subtitle: "Quant Fall 2021"
author: "Amy Henry"
date created: "10-7-2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include = FALSE}
library(tidyverse)
library(lubridate)
```

# What are the important data types in R? How does R encode and store format information about values?  

** int stands for integers.
integers are whole numbers, and function pretty much the same as doubles.

** dbl stands for doubles, or real numbers.
doubles are normal numbers with decimals, etc. The only reason there's a difference is because integers take less computer space to store; therefore some folks use them to make their code more efficient. 

** chr stands for character vectors, or strings.
Characters are "words". They're stored in quotation marks and are essentially treated as just text. 

** date stands for dates.
** dttm stands for date-times (a date + a time).
Pretty self-explanatory. Needs to be in date or date-time format for R to be able to do any kind of calculation of time. 

** lgl stands for logical, vectors that contain only TRUE or FALSE.
Logical data is yes or no, 0 or 1, TRUE or FALSE. It has only two states. We've used logical operators before to test statements. 

** fct stands for factors, which R uses to represent categorical variables with fixed possible values.
Factors include things like block, treatment, etc. and is especially useful if you plan to map or analyze based on the categorical variable. 


## Functions to test for data types and object types
Not sure what data type you're working with? Your data not behaving how you think it should when you try to graph it or compute something on it? 
Try these functions to display different characteristics of your data or objects. 

typeof()
class()
storage.mode()
length()
nrow()
ncol()
attributes()

# What do the data types look like in practice? 

This dataset contains measurements of field mice trapped in two locations over three days. 
```{r}
MouseData <- read_csv(file = "MouseDataTypes.csv")
glimpse(MouseData)
```



What data types did R automatically detect? Which ones should we fix? 

```{r}
MouseData %>% 
  mutate(
    Habitat = as.factor(Habitat), 
    MouseNumber = as.integer(MouseNumber),
    TailLength = as.numeric(TailLength),
    Weight = as.numeric(Weight),
)
```

What does it look like when a datatype is misbehaving? 
```{r}
Temp <- MouseData %>% mutate(Weight = as.character(Weight)) 
mean(Temp$Weight) # note the error message! 

mean(Temp$Female) # what did this return to us? 
```

We use functions from the package lubridate to tell R what order our date and time information comes in, and what the units are. 
```{r}
MouseData
Temp2 <- MouseData %>% select(Date, Time, DateTime) %>%
  mutate(
    Date = mdy(Date), 
    # Date = dmy(Date), # what happens if you run this line instead? 
    Time = hms(Time),
    DateTime = mdy_hm(DateTime)
)
Temp2
Temp2$DateTime[1] - Temp2$DateTime[2]
```


## All the reformatted variables together: 
```{r}
MouseData_typed <- MouseData %>% 
  mutate(
    Habitat = as.factor(Habitat), 
    Date = mdy(Date), 
    Time = hms(Time),
    MouseNumber = as.integer(MouseNumber),
    TailLength = as.numeric(TailLength),
    Weight = as.numeric(Weight),
    DateTime = mdy_hm(DateTime)
)

glimpse(MouseData_typed)
```


# What are the major types of objects in R? 

The typeof() function also helps to diagnose problems if you're struggling with a computation not doing what it should. When you apply it to more than a single vector/variable or a particular value, but to a whole data object, it can tell you the type of the object. 

#### Vector

Vectors are one-dimensional data storage in which all the data have the same data type. 

Notice how even though we entered some numbers, R decided all to store all of them as characters.
```{r}
vector_demo <- c(1, 3, "red", "applesauce")
vector_demo

typeof(vector_demo)
```

#### Matrix

```{r}
matrix_demo <- matrix(c(1:12), nrow = 3)
matrix_demo

typeof(matrix_demo)
```

#### Array

An array is like a matrix with more than two dimensions. 

```{r}
# Create two vectors of different lengths.
vector1 <- c(5,9,3)
vector2 <- c(10,11,12,13,14,15)

# Take these vectors as input to the array.
array_demo <- array(c(vector1,vector2),dim = c(3,3,2))
print(array_demo) 

class(array_demo)
```


If you'll indulge me in hurting your brain for a minute... 
```{r}
library(fivethirtyeight)

temp_Candy <- count(candy_rankings, chocolate, caramel, fruity)
temp_Candy
temp_Candy <- table(candy_rankings$chocolate, 
      candy_rankings$caramel, 
      candy_rankings$fruity)
temp_Candy
```

#### Data Frame / Tibble

Data Frames and Tibbles do exactly the same thing; the difference is only in how R handles and displays them. Tibbles are the tidyverse version, while data frames are an older format. 

```{r}
# Create two vectors of different lengths.
Var1 <- c("Adam", "Amanda", "Elliot", "Francine", "Jessie", "Caroline")
Var2 <- c(10,11,12,13,14,15)

tibble_demo <- data_frame(Var1,Var2)
tibble_demo
```

#### List

A list is a collection of data objects; each item in the list is its own object, and they don't all have to be the same length or format at all.  
```{r}
list_demo <- list(vector_demo, matrix_demo, array_demo, tibble_demo)

# selecting particular objects in the list
list_demo[[1]]
list_demo[[2]]
list_demo[[3]]
list_demo[[4]]

typeof(list_demo)
```

Technically, a tibble is a kind of list in which every item is a vector of the same length. 
```{r}
typeof(tibble_demo)
```


Often, the output of statistical tests is stored as a list.
```{r}
test_output_demo <- t.test(Var2,Var2, data = tibble_demo)
typeof(test_output_demo)
ls(test_output_demo) # if objects in a list are named, then ls() will show the names. 

test_output_demo$estimate
```

What does our final dataset look like when we assess it with different assessment metrics? What do they each do? 
```{r}
typeof(MouseData_typed)
class(MouseData_typed)
storage.mode(MouseData_typed)
length(MouseData_typed)
nrow(MouseData_typed)
ncol(MouseData_typed)
attributes(MouseData_typed)
```

What outputs from typeof() mean: 

“NULL”	NULL
“symbol”	a variable name
“pairlist”	a pairlist object (mainly internal)
“closure”	a function
“environment”	an environment
“promise”	an object used to implement lazy evaluation
“language”	an R language construct
“special”	an internal function that does not evaluate its arguments
“builtin”	an internal function that evaluates its arguments
“char”	a ‘scalar’ string object (internal only) ***
“logical”	a vector containing logical values
“integer”	a vector containing integer values
“double”	a vector containing real values
“complex”	a vector containing complex values
“character”	a vector containing character values
“…”	the special variable length argument ***
“any”	a special type that matches all types: there are no objects of this type
“expression”	an expression object
“list”	a list
“bytecode”	byte code (internal only) ***
“externalptr”	an external pointer object
“weakref”	a weak reference object
“raw”	a vector containing bytes
“S4”	an S4 object which is not a simple object


